A Linear Solution of the Linearity Challenge

### 2025-12-26 paper/Congruence.lagda.tex
revision

Here the process
$(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, that is
$\Select\y\LeftTag\z.P$, is found under a cut for $x : A$ and is using some
channel $y : B \Plus C$ to send the tag $\InTag_1$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ uses a fresh continuation channel $z : B$.
Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type of the continuation $z$ comes before the type of the restricted channel
$x$.
%
After structural precongruence is applied, however, the type $B$ of the
continuation channel $z$ ends up behind that of the restricted channel $x$,
because now the two channels are introduced in the opposite order. Therefore, we
need to rename the channels in $P$ so that it is well typed in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
renaming is achieved applying the function \AgdaFunction{↭proc} to the
\AgdaInductiveConstructor{swap} permutation and to the process $P$.

@@

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, meaning
that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
\AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$
is replaced by $\AgdaInductiveConstructor{>}~r$.

>>>>

Here the process
$(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, that is
$\Select\y\LeftTag\z.P$, is found under a cut for $x : A$ and is using some
channel $y : B \Plus C$ to send the tag $\InTag_1$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ uses a fresh continuation channel $z : B$.
Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type $B$ of the continuation $z$ comes \emph{before} the type $A$ of the
restricted channel $x$.
%
After structural pre-congruence is applied, however, the type of the
continuation channel $z$ ends up behind that of the restricted channel $x$,
because now $z$ and $x$ are introduced in the opposite order. Therefore, we need
to rename the channels in $P$ so that it is well typed in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
renaming is achieved applying the function \AgdaFunction{↭proc} to the
\AgdaInductiveConstructor{swap} permutation and to the process $P$.

@@

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, implying
that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, which is similar to
\AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$
is replaced by $\AgdaInductiveConstructor{>}~r$.

### 2025-12-26 paper/DeadlockFreedom.lagda.tex
revision

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalise in Agda. First of all we need to define the notion of
\emph{thread}, that is any process other than a cut. It is convenient to provide
a more fine-grained classification of threads, distinguishing between links,
input/output actions, and sometimes also whether such actions operate on free or
bound channels. We define predicates for each of these process classes:

@@

These predicates are relatively uninteresting since they are essentially
isomorphic to the relevant fragments of the \AgdaDatatype{Proc} datatype. The
only aspect that is worth pointing out here is that in \AgdaDatatype{Input},
\AgdaDatatype{Output} and \AgdaDatatype{Server} the channel being acted upon by
the thread is the \emph{first} in $\Context$, hence it is the \emph{most
recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
those threads that, in the context of a cut, operate on the channel bound by the
cut or on a free channel. To clarify this aspect, consider the
\AgdaInductiveConstructor{wait} constructor in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

@@

\begin{code}
  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (ch ⟨ < p ⟩ P))
\end{code}
where the use of the \AgdaInductiveConstructor{<} constructor indicates that the
thread operates on the most recent channel. In the latter predicate we have

@@

\begin{code}
  wait     : ∀{C Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Delayed (wait (ch ⟨ >_ {C} p ⟩ P))
\end{code}
where the use of the \AgdaInductiveConstructor{>} constructor indicates that the
thread operates on a channel other than the most recent one. Note that here we
have to specify the type $C$ in front of $\Context$ or else Agda is unable to
automatically resolve some metavariables.

@@

Now the \AgdaDatatype{Thread} predicate is simply the disjoint union of all the
previous ones.

@@

We define observability, reducibility and aliveness in the expected way:

@@

In order to prove that every (well-typed) process is alive, it is convenient to
also define a ``canonical'' form for cuts, that is a form that matches at least
one of the left-hand-sides of the rules for structural pre-congruence or
reduction (\Cref{tab:semantics}). This is the notion where the fine-grained
classification of threads introduced earlier comes into play.

>>>>

As we have seen in \Cref{sec:calculus}, the deadlock freedom property and
\Cref{thm:df} rest on some notions and predicates about processes which must be
formalised in Agda. First of all we need to define the notion of \emph{thread},
that is any process other than a cut. It is convenient to provide a more
fine-grained classification of threads, distinguishing between links and
input/output actions and sometimes also on whether such actions operate on free
or bound channels. We define predicates for each of these classes:

@@

The implementation of these predicates is not interesting since it is
essentially isomorphic to the relevant fragments of the \AgdaDatatype{Proc}
datatype. The only aspect that is worth pointing out here is that in
\AgdaDatatype{Input}, \AgdaDatatype{Output} and \AgdaDatatype{Server} the
channel being acted upon by the thread is the \emph{first} in $\Context$, hence
it is the \emph{most recently introduced} channel, whereas in
\AgdaDatatype{Delayed} and \AgdaDatatype{DelayedServer} the channel is not the
first. This allows us to distinguish those threads that, in the context of a
cut, operate on the channel bound by the cut or on a free channel. To clarify
this aspect, we show the implementation of the constructor
\AgdaInductiveConstructor{wait} in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

@@

\begin{code}
  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (ch ⟨ < p ⟩ P))
\end{code}
where the use of the constructor \AgdaInductiveConstructor{<} indicates that the
thread operates on the most recent channel. In the latter predicate we have

@@

\begin{code}
  wait     : ∀{C Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Delayed (wait (ch ⟨ >_ {C} p ⟩ P))
\end{code}
where the use of the constructor \AgdaInductiveConstructor{>} indicates that the
thread operates on a channel other than the most recent one. Note that here we
have to specify the type $C$ in front of $\Context$ or else Agda is unable to
automatically resolve some metavariables.

@@

The predicate \AgdaDatatype{Thread} is simply the disjoint union of all the
previous ones.

@@

Observability, reducibility and aliveness are defined in the expected way:

@@

In order to prove that every (well-typed) process is alive, it is convenient to
define a ``canonical'' form for cuts, that is a form that matches at least one
of the l.h.s of one of the rules for structural pre-congruence or reduction in
\Cref{tab:semantics}. This is the notion where the fine-grained classification
of threads introduced earlier comes into play.

### 2025-12-26 paper/Reduction.lagda.tex
revision

Just like structural precongruence, reduction is formalised as a binary relation
between processes that are well typed in the very same typing context. Thus, the
definition of reduction embeds the subject reduction result.

@@

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
only free channel occurring in a process of the form $\Close\x$, the
\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
$(\AgdaInductiveConstructor{<}~q_0)$ to be sure that the referenced channel is
also the restricted one, and therefore matches the one of the
\AgdaInductiveConstructor{close} process.

>>>>

Just like structural pre-congruence, reduction is formalised as a binary
relation between processes that are well typed in the same typing context. Thus,
the definition of reduction embeds subject reduction (\Cref{thm:red}).

@@

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•}$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ is on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

While the constructor \AgdaInductiveConstructor{close} implicitly refers to the
only free channel occurring in a process of the form $\Close\x$, the constructor
\AgdaInductiveConstructor{wait} uses a splitting proof of the form
$\AgdaInductiveConstructor{<}~q_0$ to make sure that the referenced channel is
also the restricted one, and therefore matches the one of the
\AgdaInductiveConstructor{close} process.
